Modelling the City of Melbourne Pedestrian Data

### 2017-10-01 presentation/final.Rmd
update

ggplot(original_df %>%  filter(Date_Time < "2014-07-06 00:00:00",
                               Date_Time > "2014-07-04 12:00:00")) +
  geom_line(aes(x = Date_Time, y = fit_dat_plot),
            colour = plasma(1, begin = .75, end = 0),
            size = 2) +
  geom_line(aes(x = Date_Time, y = `Southern Cross Station`),
            size  = 2) +
  theme_minimal() + xlab("Time") +
  ylab("Pedestrian Counts") +
  ggtitle("Pedestrian Counts at Southern Cross Station on 04/07/2014 - 05/07/2014")
```

@@

sens_lines <- data.frame(lon = sens_lines_lon,
                         lat = sens_lines_lat)
ggmap(melb3) + geom_line(aes(x = lon, y = lat),
                        data = sens_lines, size = 2.5,
                        colour = viridis(3)[2], alpha = 0.5) +
               geom_point(aes(x = lons, y = lats, colour = cl),
                          size = 8,
                          data = loc) +
               scale_colour_viridis(discrete = TRUE,
                                    name = "Sensor NA Prop.",
                                    begin = 0.5, end = 0) +
               geom_point(aes(x = loc["Southbank",]$lons,
                              y = loc["Southbank",]$lats),
                          size = 16, colour = viridis(3)[2]) +
               geom_text(aes(label = ids, x = lons, y = lats),
                         data = loc, size = 4,
                         nudge_y = -0.0002) +
              xlab(NULL) + ylab(NULL) +
              theme(plot.title = element_text(hjust = 0.5),
                    plot.caption = element_text(hjust = 0.5),
                    axis.text.x = element_blank(),
                    axis.text.y = element_blank(),
                    axis.ticks = element_blank(),
                    legend.position = c(.8,0.2))

@@

p1 <- ggplot(tr_dat %>% filter(Date_Time < "2017-03-02",
                         Date_Time > "2017-02-26")) +
      geom_line(aes(x = Date_Time, y = Count,
                    colour = Sensor), size = 2) +
      scale_colour_viridis(discrete = TRUE,
                           labels = c("Neighbour 1 (Flinders St Station Underpass)",
                                      "Neighbour 2 (Sandridge Bridge)",
                                      "Target (Southbank)")) +
      xlab("Time") + theme_minimal() + ggtitle("Unscaled Neighbours") +
      theme(legend.position = "none")

p2 <- ggplot(tr_dat_sc %>% filter(Date_Time < "2017-03-02",
                         Date_Time > "2017-02-26")) +
      geom_line(aes(x = Date_Time, y = Count,
                    colour = Sensor), size = 2) +
      scale_colour_viridis(discrete = TRUE,
                           labels = c("Neighbour 1 (Flinders St Station Underpass)",
                                      "Neighbour 2 (Sandridge Bridge)",
                                      "Target (Southbank)")) +
      xlab("Time") + ggtitle("Scaled/Stadardised Neighbours") +
      ylab("Standard Deviation") + theme_minimal() +
      theme(legend.position = "bottom")
```

@@

- general code which will run the algorithm
- inputs required: 
  * data in wide format
  * threshold value for missing values proportion (_default:_ `threshold = 0.1`)
  * max length of repeated values (_default:_ `maxlength = 17`)
- use R and RStudio with packages:
  * `dplyr`, `lubridate` and `tidyr` for data manipulation
  * `foreach` and `doSNOW` for parallel computing
  * `rwalkr` for accessing latest data
---
class: center
## Imputation Model Evaluation
### In-sample fit

@@

```{r, echo = FALSE, fig.retina = 6, message=FALSE, warning=F, fig.width = 4.5, fig.height = 3.5}
ggplot(test_df, aes(x = `Bourke Street Mall (North)`,
                    y = test_fit$`Bourke Street Mall (North)`)) +
                  geom_point(alpha = 0.5) +
                  scale_x_continuous(limits = c(0, 5000)) +
                  scale_y_continuous(limits = c(0, 5000)) +
                  xlab("Actual") + ylab("Fitted") + theme(aspect.ratio = 1) +
                  geom_abline(aes(slope = 1, intercept = 0),
                              colour = viridis(3)[1],
                              alpha = 0.7, size = 1.5,
                              linetype = 2) +
                  geom_smooth(method = "lm", size = 1.5, 
                       colour = viridis(3)[3]) +
                  ggtitle("Actual vs Fitted \n at Bourke Street Mall (North) in 2017")

@@

# Predictive model using imputed data
- can perform analysis at all locations now with complete data
```{r, echo = FALSE, fig.retina = 6, message=FALSE, warning=F, fig.width = 10, fig.height = 3}
melb_cent_mon9 <- dfa2 %>%
  filter(Day == "Monday",
         Time == 9) %>%
         .$`Melbourne Central`

@@

p1 <- ggplot() + geom_line(aes(x = 1:length(melb_cent_mon9),
                             y = melb_cent_mon9)) +
      theme_minimal() + xlab("Time (nth Week since 2014)") +
      geom_point(aes(x = 1:length(melb_cent_mon9),
                     y = 0, alpha = monday_hdays),
                 size = 3) + 
      scale_alpha_discrete(name = "Is public holiday") + 
      ylab("Counts at Monday 9am")
p2 <- ggplot() + geom_density(aes(x = melb_cent_mon9)) +
                 theme_minimal() + xlab("Counts at Monday 9am")

@@

```
- can treat each specific hour of the week as a time series
  * find covariates outside the counts data to help predict counts
  * eg. weather data, public events

### Objective: On the fly predictions of pedestrian traffic
- requires efficient data storage

---
# Future steps for final paper
- develop Shiny app wih RStudio to provide on the fly predictions
  * make the predictive models more accessible
  * provide visualisations to present the predictions

```{r, cache = TRUE, fig.retina = 4, echo = FALSE, message = FALSE, fig.align= 'center', fig.width=5, fig.height=4, warning=F}
mar13_16_8am <- ped_data[ped_data$Date == "2016-03-13", ] %>%
                filter(Time == 8)

@@

>>>>

ggplot(original_df %>%  filter(Date_Time < "2014-07-06 00:00:00",
                               Date_Time > "2014-07-04 12:00:00")) +
  geom_line(aes(x = Date_Time, y = fit_dat_plot),
            colour = plasma(1, begin = .75, end = 0),
            size = 2) +
  geom_line(aes(x = Date_Time, y = `Southern Cross Station`),
            size  = 2) +
  theme_minimal() + xlab("Time") +
  ylab("Pedestrian Counts") +
  ggtitle("Pedestrian Counts at Southern Cross Station \n on 04/07/2014 - 05/07/2014") +
  theme(plot.title = element_text(hjust = 0.5))
```

@@

sens_lines <- data.frame(lon = sens_lines_lon,
                         lat = sens_lines_lat)
ggmap(melb3) + geom_line(aes(x = lon, y = lat),
                        data = sens_lines, size = 2.5,
                        colour = viridis(3)[2], alpha = 0.5) +
               geom_point(aes(x = lons, y = lats, colour = cl),
                          size = 8,
                          data = loc) +
               scale_colour_viridis(discrete = TRUE,
                                    name = "Sensor NA Prop.",
                                    begin = 0.5, end = 0) +
               geom_point(aes(x = loc["Southbank",]$lons,
                              y = loc["Southbank",]$lats),
                          size = 16, colour = viridis(3)[2]) +
               geom_text(aes(label = ids, x = lons, y = lats),
                         data = loc, size = 4,
                         nudge_y = -0.0003) +
              xlab(NULL) + ylab(NULL) +
              theme(plot.title = element_text(hjust = 0.5),
                    plot.caption = element_text(hjust = 0.5),
                    axis.text.x = element_blank(),
                    axis.text.y = element_blank(),
                    axis.ticks = element_blank(),
                    legend.position = c(.8,0.2))

@@

p1 <- ggplot(tr_dat %>% filter(Date_Time < "2017-03-02",
                         Date_Time > "2017-02-26")) +
      geom_line(aes(x = Date_Time, y = Count,
                    colour = Sensor), size = 2) +
      scale_colour_viridis(discrete = TRUE,
                           labels = c("Neighbour 1 (Flinders St Station Underpass)",
                                      "Neighbour 2 (Sandridge Bridge)",
                                      "Target (Southbank)")) +
      xlab("Time") + theme_minimal() + ggtitle("Unscaled Neighbours") +
      theme(legend.position = "none",
            plot.title = element_text(hjust = 0.5))

p2 <- ggplot(tr_dat_sc %>% filter(Date_Time < "2017-03-02",
                         Date_Time > "2017-02-26")) +
      geom_line(aes(x = Date_Time, y = Count,
                    colour = Sensor), size = 2) +
      scale_colour_viridis(discrete = TRUE,
                           labels = c("Neighbour 1 (Flinders St Station Underpass)",
                                      "Neighbour 2 (Sandridge Bridge)",
                                      "Target (Southbank)")) +
      xlab("Time") + ggtitle("Scaled/Stadardised Neighbours") +
      ylab("Standard Deviation") + theme_minimal() +
      theme(legend.position = "bottom",
            plot.title = element_text(hjust = 0.5))
```

@@

- general code which will run the algorithm
- inputs required: 
  * data in wide format
  * threshold value for missing values proportion (_default:_ `threshold = 0.1`)
  * max length of repeated values (_default:_ `maxlength = 17`)
  * `threshold` and `maxlength` are tuning parameters
- use R and RStudio with packages:
  * `dplyr`, `lubridate` and `tidyr` for data manipulation
  * `foreach` and `doSNOW` for parallel computing
  * `rwalkr` for accessing latest data
---
class: center
## Imputation Model Evaluation
### In-sample fit

@@

```{r, echo = FALSE, fig.retina = 6, message=FALSE, warning=F, fig.width = 4.5, fig.height = 3.5}
ggplot(test_df, aes(x = `Bourke Street Mall (North)`,
                    y = test_fit$`Bourke Street Mall (North)`)) +
                  geom_point(alpha = 0.2) +
                  scale_x_continuous(limits = c(0, 5000)) +
                  scale_y_continuous(limits = c(0, 5000)) +
                  xlab("Actual") + ylab("Fitted") + theme(aspect.ratio = 1) +
                  geom_abline(aes(slope = 1, intercept = 0),
                              colour = viridis(3)[1],
                              alpha = 0.7, size = 1.5,
                              linetype = 2) +
                  geom_smooth(method = "lm", size = 1.5, 
                       colour = viridis(3)[3], alpha = 0.7) +
                  ggtitle("Actual vs Fitted \n at Bourke Street Mall (North) in 2017")

@@

## Predictive model using imputed data
- can perform analysis at all locations now with complete data
```{r, echo = FALSE, fig.retina = 6, message=FALSE, warning=F, fig.width = 10, fig.height = 2.5}
melb_cent_mon9 <- dfa2 %>%
  filter(Day == "Monday",
         Time == 9) %>%
         .$`Melbourne Central`

@@

p1 <- ggplot() + geom_line(aes(x = 1:length(melb_cent_mon9),
                             y = melb_cent_mon9)) +
      theme_minimal() + xlab("Time (nth Week since 2014)") +
      geom_point(aes(x = 1:length(melb_cent_mon9),
                     y = 0, alpha = monday_hdays),
                 size = 3) + 
      scale_alpha_discrete(name = "Is public holiday",
                           range = c(0.01,1)) + 
      ylab("Counts at Monday 9am")
p2 <- ggplot() + geom_density(aes(x = melb_cent_mon9)) +
                 theme_minimal() + xlab("Counts at Monday 9am")

@@

```
- can treat each specific hour of the week as a time series
  * find covariates outside the counts data to help predict counts
  * eg. weather data, public events
- can use clustering methods to investigate similarities between locations
  * potentially use different models based on cluster membership
  * eg. train stations, parks and recreation, shopping
#### Objective: On the fly predictions of pedestrian traffic
- requires efficient data storage

---
# Future steps for final paper
- perform simulations to test imputation algorithm
- develop Shiny app wih RStudio to provide on the fly predictions
  * make the predictive models more accessible
  * provide visualisations to present the predictions

```{r, cache = TRUE, fig.retina = 4, echo = FALSE, message = FALSE, fig.align= 'center', fig.width=5, fig.height=3.5, warning=F}
mar13_16_8am <- ped_data[ped_data$Date == "2016-03-13", ] %>%
                filter(Time == 8)

@@

---
# Summary
- investigated issues with the Melbourne CBD pedestrian data
- proposed algorithm for imputation of missing values in pedestrian data using split approach based on missing values proportion
-

---
# Acknowledgements

References:
- City of Melbourne Open Data Portal - https://data.melbourne.vic.gov.au/
- An Introduction to Statistical Learning - _G. James, D. Witten, T. Hastie and R. Tibshirani_
-

R packages used:
  * `dplyr`, `lubridate` and `tidyr` for data manipulation
  * `foreach` and `doSNOW` for parallel computing
  * `rwalkr` for accessing latest data (Thanks Earo!)
  * `ggplot2` and `ggmap` for visualisations
  * `xaringan` for compliling this presentation

GitHub repository for this project: https://github.com/gavinchin/honours2017
